Galaxy Imaging with Generative Models: Insights from a Two-Models Framework

### 2025-03-12 MNRAS/galaxy-generative-models-overleaf-submit-2nd/article_MNRAS.tex
corr. Section 2

The diffusion model shows a transition from memorization to generalization as the dataset size increases, confirming previous findings by \cite{kadkhodaie2024generalization}. Smaller datasets lead to overfitting, while larger datasets enable novel sample generation, supported by the denoising process's theoretical basis. For the flow-based model, we propose an "inversion test" leveraging its bijective nature. Similarly, the GAN-based model achieves comparable morphological consistency but lacks bijectivity. We then introduce a "discriminator test," which shows successful learning for larger datasets but poorer confidence with smaller ones. Across all models, dataset sizes below $O(10^5)$ pose challenges to learning.

Along our experiments, the "two-models" framework introduced by \cite{kadkhodaie2024generalization} enables robust evaluations, highlighting both the potential and limitations of these models. These findings provide valuable insights into statistical learning in generative modeling, with applications certainly extending beyond galaxy image generation.
\end{abstract}

@@

To optimize the generator, a second network ($\bm{D}$), acting as a \textit{discriminator}, aims to determine whether a sample $\bm{x}$ is from the model distribution or the dataset distribution. To achieve this, one solves the \texttt{min-max} problem:
\begin{equation}
\min_{\bm{G}} \max_{\bm{D}}\left\{ \Esp_{\bm{x}\sim p_{data}(\bm{x})}[\log \bm{D}(\bm{x})] + \Esp_{\bm{z}\sim \pi(\bm{z})}[\log(1-\bm{D}(\bm{G}(\bm{z})))] \right\},
\end{equation}
where $p_{data}$ is the data-generating distribution. Both $\bm{G}$ and $\bm{D}$ are parameterized networks. Note that, at fixed parameters of $\bm{G}$, the optimal \textit{discriminator} $\bm{D}^\ast$ has the following expression:
\begin{equation}
\bm{D}^\ast(\bm{x}) = \frac{p_{data}(\bm{x})}{p_{data}(\bm{x}) + p_{\bm{G}}(\bm{x})},
\end{equation}
with $p_{\bm{G}}(\bm{x}) = \pi(\bm{z})|\det J_{\bm{G}}|^{-1}$, where $J_{\bm{G}}$ is the Jacobian of the \textit{generator}. This leads to a reformulation of the \texttt{min-max} problem as follows:
\begin{equation}
\min_G \left\{ \DKL{\left( p_{data} \left\| \frac{p_{data} + p_{\bm{G}}}{2}\right. \right)} + \DKL{\left(p_{\bm{G}} \left\| \frac{p_{data} + p_{\bm{G}}}{2} \right. \right)} \right\},
\end{equation}
yielding the optimal solution $p_{\bm{G}} = p_{data}$, which perfectly matches the dataset distribution.

@@

GAN architectures have been used for galaxy generation in conjunction with deblending and deconvolution tasks. For instance, \cite{Schawinski2017} and \cite{Hemmati_2022} employed vanilla-GANs, while \cite{Coccomini2021} utilized the \texttt{light-weight-GAN} that we use in our experiment too.

\subsection{Normalizing Flows}
\label{sec-NF}
%
Looking at Equation \ref{eq-VAE-ELBO}, one notices that the optimal solution is reached when $\DKL(p_{\bm{\phi}}(\bm{z}|\bm{x})\| p(\bm{z}|\bm{x})) = 0$, which is true if and only if $p_{\bm{\phi}}(\bm{z}|\bm{x}) = p(\bm{z}|\bm{x})$. However, this situation is generally unattainable because typical $p_{\bm{\phi}}$ parameterizations involve, for instance, independent Gaussian distributions. This limitation in parameterization is one of the drawbacks of VAEs. This motivates the use of normalizing flows (NF) to introduce more flexibility in the family of distributions, with the hope that the true posterior is a member of such a family \citep{Tabak2010, Tabak2013a, Rezende2015}.

A flow $T$ is defined as a diffeomorphism (i.e., a bijector) between the data space and a latent space, such that:
\begin{align}
\bm{x} = T(\bm{z}) & \quad \Leftrightarrow \quad \bm{z} = T^{-1}(\bm{x}).
\end{align}
If the distribution of $\bm{z}$ is $\pi(\bm{z})$ (e.g., $\mathcal{N}(\bm{0},\bm{1})$), then the following relation applies:
\begin{equation}
p(\bm{x}) = \pi(\bm{z})|\det J_T(\bm{z})|^{-1} = \pi(T^{-1}(\bm{x}))|\det J_{T^{-1}}(\bm{x})|,
\end{equation}
where $J_T$ and $J_{T^{-1}}$ are the Jacobians of the transformations $T$ and $T^{-1}$, respectively. Unlike VAEs and GANs, the dimensionalities of the latent and data spaces are the same by definition in flow-based models.

The flow $T$ is generally constructed as a composition of several individual flows $\{T_i\}_{i<n}$ such that:
\begin{align}
T = T_1 \circ T_2 \circ \dots \circ T_n & \quad \Leftrightarrow \quad 
T^{-1} = T_n^{-1} \circ T_{n-1}^{-1} \circ \dots \circ T_1^{-1}.
\end{align}
Figure \ref{fig-normflow} illustrates a schematic view of the \textit{forward} or \textit{generative} direction and the \textit{backward/reverse} or \textit{training} direction. If we denote $\bm{z}_i = T_i(\bm{z}_{i-1})$ for all $i < n$ (using $\bm{z}_0 = \bm{z}$ and $\bm{z}_n = \bm{x}$), then:
\begin{equation}
\log |\det J_T| = \sum_{i=0}^{n-1} \log |\det J_{T_i}(\bm{z}_{i-1})|.
\label{eq-flow-jacob}
\end{equation}

@@

Let the flow be parametrized by a vector $\bm{\theta}$, leading to a parametrized flow model $p_{\bm{\theta}}(\bm{x})$. To optimize the flow, we consider the $\DKL$ divergence between $p_{\bm{\theta}}(\bm{x})$ and the true $p(\bm{x})$, resulting in\footnote{For simplicity, we assume $\pi(\bm{z}) = \mathcal{N}(\bm{0}, \bm{1})$ without introducing additional unknown parameters.}:
\begin{align}
\mathcal{L}(\bm{\theta}) &= \DKL(p(\bm{x})\|p_{\bm{\theta}}(\bm{x})) 
= - \Esp_{\bm{x}\sim p(\bm{x})}[-\log p_{\bm{\theta}}(\bm{x})] + \mathrm{const.} \nn \\
&= - \Esp_{\bm{x}\sim p(\bm{x})}[\log \pi(T_{\bm{\theta}}^{-1}(\bm{x})) + \log |\det J_{T_{\bm{\theta}}^{-1}}(\bm{x})|] + \mathrm{const.}
\end{align}

In practice, computing the expectation $\Esp_{\bm{x} \sim p(\bm{x})}$ is not feasible and is replaced by a Monte Carlo approach using the data samples $\{\bm{x}^{(i)}\}_{i<N}$. Dropping the constant term, the loss function becomes:
\begin{equation}
\mathcal{L}(\bm{\theta}) = -\frac{1}{N} \sum_{i=0}^{N-1} \left\{ \log \pi(T_{\bm{\theta}}^{-1}(\bm{x}^{(i)})) + \log |\det J_{T_{\bm{\theta}}^{-1}}(\bm{x}^{(i)})| \right\}.
\end{equation}

The computation of the determinant of the Jacobian and its gradient can be challenging. However, it becomes more straightforward if the Jacobian matrices $\{J_{T_i}^{-1}\}_{i<n}$ (Eq.~\ref{eq-flow-jacob}) are triangular. To minimize $\mathcal{L}(\bm{\theta})$ and obtain the best $\bm{\theta}$, we require $T_\theta^{-1}$ as well as its Jacobian and gradients. Additionally, $T_\theta$ is needed to generate new samples $\bm{x}$ from latent samples $\bm{z} \sim \pi(\bm{z})$. Various architectures of normalizing flows, such as \textit{autoregressive flows}, enable such operations, as reviewed in \cite{Papamakarios2021}.

The general schema of \textit{autoregressive flows} is as follows: let $\bm{x}, \bm{z} \in \mathbb{R}^d$, and denote $\bm{x} = (x_1, x_2, \dots, x_d)$ and $\bm{z} = (z_1, z_2, \dots, z_d)$. Then:
\begin{equation}
\forall i, \quad x_i = T(z_i; c_i) \quad \mathrm{with} \quad c_i = C_i(z_1, \dots, z_{i-1}) \triangleq C_i(z_{<i}),
\end{equation}
where $T$ is a strictly increasing monotonic transformation, and $C_i$ is known as the \textit{transformer}\footnote{This term should not be confused with transformer networks used in deep learning architectures based on multi-head attention mechanisms \citep{Vaswani2017}.} and the \textit{conditioner}.

The properties of such flows ensure the invertibility of the \textit{transformer} and the triangular structure of the Jacobian:
\begin{equation}
J_T = \left[ \frac{\partial x_i}{\partial z_j} \right] = 
\begin{cases}
0 & \mathrm{for}\ i > j, \\
\frac{\partial x_i}{\partial z_i}(z_i; c_i) & \mathrm{for}\ i = j.
\end{cases}
\end{equation}
This leads to:
\begin{equation}
\log |\det J_T| = \sum_{i=1}^d \log \left| \frac{\partial x_i}{\partial z_i}(z_i; c_i) \right|.
\end{equation}
Different types of transformers and conditioners yield distinct architectures.

Among the transformers, the class of affine transformations is perhaps the simplest. These are defined as:
\begin{equation}
T(z_i; c_i) = e^{\alpha_i} z_i + \beta_i, \quad \mathrm{with} \quad c_i = (\alpha_i(z_{<i}), \beta_i(z_{<i})).
\label{eq-affine-coupling}
\end{equation}
Using matrix notation, this can be expressed as $T(\bm{z}) = \exp(\bm{\alpha}) \odot \bm{z} + \bm{\beta}$, where $\odot$ denotes the Hadamard product. The invertibility of $T$ is guaranteed by the exponential function, and the Jacobian determinant simplifies to the $e^{\alpha_i}$ factor. Affine transformers are used in works such as \citep{DinhKB14, Papamakarios2017a, DinhSB17, Kingma2018}. 
Furthermore, \textit{spline}-based flows have gained traction, as exemplified in \cite{Crenshaw_2024}, where the \texttt{PZFlow} code is applied to model galaxy photometric redshift posterior distributions.

Concerning the conditioners $C_i$, in principle, they can be any functions (or models) that take $z_{<i}$ as input and output $c_i$. However, practical implementations are driven by computational cost. A particular approach is the \textit{coupling layer}, implemented in \texttt{Real NVP}\footnote{n.b stands for \textit{real-valued non-volume preserving} transformation.} as proposed in \citep{DinhKB14, DinhSB17}. For an affine transformation $T$ (Eq.~\ref{eq-affine-coupling}) with $s < d$ (a common choice is $s = d/2$), the forward pass is defined as:
\begin{equation}
(\mathrm{NVP}) \quad \begin{cases}
x_{\leq s} &= z_{\leq s}, \\
x_{s+1,\dots,d} &= z_{s+1,\dots,d} \odot \exp(\bm{\alpha}(z_{\leq s})) + \bm{\beta}(z_{\leq s}),
\end{cases}
\label{eq-NVP_forward}
\end{equation}
where $\bm{\alpha}(z_{\leq s})$ and $\bm{\beta}(z_{\leq s})$ are functions of the first $s$ components of $\bm{z}$.

The backward pass, involving the inverse transformation, has the same computational complexity—unlike other conditioners such as the masked schema used in \citep{Germain2015}—and is given by:
\begin{equation}
(\mathrm{NVP}^{-1}) \quad \begin{cases}
z_{\leq s} &= x_{\leq s}, \\
z_{s+1,\dots,d} &= (x_{s+1,\dots,d} - \bm{\beta}(x_{\leq s})) \odot \exp(-\bm{\alpha}(x_{\leq s})).
\end{cases}
\label{eq-NVP_backward}
\end{equation}

The Jacobian matrix in this schema is triangular, with the diagonal given by:
\begin{equation}
\diag(J_T) = (\bm{1}_s, \diag(\exp(\bm{\alpha}(z_{\leq s})))),
\end{equation}
where $\bm{1}_s$ is an $s \times s$ identity matrix. This leads to $d-s$ non-unity diagonal terms, reflecting the \textit{non-volume conservation} property of the transformation (hence the "NVP" notation).

The structure of the transformation (e.g., Equation~\ref{eq-NVP_forward}) involves a single split of the $\bm{z}$ vector coordinates into two parts. The first part, $z_{\leq s}$, is left untransformed, while the transformation of the second part, $z_{>s}$, depends only on $z_{\leq s}$. Notably, the conditioner functions $\bm{\alpha}$ and $\bm{\beta}$ depend exclusively on $z_{\leq s}$ for both the forward and backward transformations. This single-split schema is illustrated in Figure~\ref{fig-nvp}.
%
\begin{figure}
\centering
\includegraphics[width=\linewidth]{fig-nvp.pdf}
\includegraphics[width=\linewidth]{fig-nvp-inverse.pdf}
\caption{Affine coupling layer schema of forward and backward directions (Equations~\ref{eq-NVP_forward},\ref{eq-NVP_backward}) with a single split mechanism (figure inspired by \protect\cite{Papamakarios2021}).}
\label{fig-nvp}
\end{figure}
%
This splitting schema, implemented in a so-called \textit{coupling layer}, makes the flow-based model computationally efficient. By using deep neural networks, one can achieve complex $(\bm{\alpha},\bm{\beta})$ functions. Designers of complete architectures can compose different \textit{coupling layers}, alternating the order of $\bm{z}$ elements between layers to ensure that all elements are transformed by the end of the sequence of operations. A generalization of this permutation can be achieved using a $1\times 1$ convolution layer.

\cite{DinhSB17} improved the modeling further by introducing a multi-scale approach, which is too extensive to detail here. \cite{Kingma2018} utilized these extensions to build the \texttt{Glow} model, which is employed in our experiments. For completeness, the schema of this model is depicted in Figure~\ref{fig-Glow-archi} bored from Figure 2 of \cite{Kingma2018}. In the \texttt{Glow} model, the prior is learned during the optimization of the flows, but to generate new samples, one still uses $\bm{z} \sim \mathcal{N}(\bm{0}, \bm{1})$, where $\bm{z}$ comprises tensors corresponding to the number of levels $L$.

@@

Using the presentation framework of \cite{kadkhodaie2024generalization}, the transport process in the \textit{score-based diffusion} algorithm is modeled by the Ornstein-Uhlenbeck equation \citep{Uhlenbeck1930}. The \textit{forward} process follows this stochastic differential equation:
\begin{equation}
d\bm{x}_t = -\bm{x}_t \, dt + \sqrt{2}\, dB_t \quad t\in[0,T],
\label{eq-Ornstein-cont}
\end{equation}
where $dB_t$ represents a Wiener process (Brownian motion) and $T$ is a finite endpoint assumed to be sufficiently large. The solution to this forward process is:
\begin{equation}
\bm{x}_t = \bm{x}_0 e^{-t} + \underbrace{(1-e^{-2t})^{1/2}}_{\sigma_t} \bm{z}, \qquad \bm{z} \overset{\text{iid}}{\sim} \mathcal{N}(\bm{0},\bm{1}).
\end{equation}
The corresponding probability density function is expressed as:
\begin{equation}
p_t(\bm{x}_t) = \int p_t(\bm{x}_t,\bm{x}_0) \, d\bm{x}_0 = \int p_t(\bm{x}_t|\bm{x}_0) p_0(\bm{x}_0) \, dx_0,
\label{eq-smoothing-p0-gauss}
\end{equation}
where, based on the Fokker-Planck equation for the Ornstein-Uhlenbeck process:
\begin{equation}
p_t(\bm{x}_t|\bm{x}_0) = \frac{1}{(2\pi \sigma_t^2)^{d/2}} \exp\left\{-\frac{\|\bm{x}_t-\bm{x}_0 e^{-t}\|^2}{2\sigma_t^2}\right\}.
\end{equation}
Thus, $p_t(\bm{x}_t)$ is the convolution of the initial distribution $p_0(\bm{x}_0)$ with a Gaussian of variance $\sigma_t^2$. This gradually blurs the distribution, leading $p_t(\bm{x}_t)$ to converge towards the normal distribution $\mathcal{N}(\bm{0},\bm{1})$.

@@

The central challenge in generation lies in estimating the \textit{score}. Two main approaches have been proposed: (1) deriving a model for $p_t$ that incorporates regularity patterns and correlations to learn the \textit{score} via \textit{score matching} \citep{hyvarinen2005a}, or (2) employing a \textit{denoising} neural network, as introduced by \cite{Bengio2013}, further developed in \cite{Sohl-Dickstein2015,Ho2020}, and recently examined from a mathematical perspective in \cite{kadkhodaie2024generalization}. The first approach has been utilized in works such as \cite{Guth2022b,Lempereur2024} with Gibbs energy parameterization. The second has found applications in astrophysics, including galaxy image generation \citep{smith2021}, generation of 21~cm luminosity temperature maps \citep{Zhao2023}, super-resolution of large-scale cosmic structures \citep{Schanz2023}, and Bayesian posterior sampling for weak lensing mass-mapping problems \citep{Remy2023}.

>>>>

The diffusion model shows a transition from memorization to generalization as the dataset size increases, confirming previous findings. Smaller datasets lead to overfitting, while larger datasets enable novel sample generation, supported by the denoising process's theoretical basis. For the flow-based model, we propose an "inversion test" leveraging its bijective nature. Similarly, the GAN-based model achieves comparable morphological consistency but lacks bijectivity. We then introduce a "discriminator test," which shows successful learning for larger datasets but poorer confidence with smaller ones. Across all models, dataset sizes below $O(10^5)$ pose challenges to learning.

Along our experiments, the "two-models" framework enables robust evaluations, highlighting both the potential and limitations of these models. These findings provide valuable insights into statistical learning in generative modeling, with applications certainly extending beyond galaxy image generation.
\end{abstract}

@@

To optimize the generator, a second network ($\bm{D}$), acting as a \textit{discriminator}, aims to determine whether a sample $\bm{x}$ is from the model distribution or the dataset distribution. To achieve this, one solves the \texttt{min-max} problem ({\color{red} Nash equilibrium}):
\begin{equation}
\min_{\bm{G}} \max_{\bm{D}}\left\{ \Esp_{\bm{x}\sim p_{data}(\bm{x})}[\log \bm{D}(\bm{x})] + \Esp_{\bm{z}\sim \pi(\bm{z})}[\log(1-\bm{D}(\bm{G}(\bm{z})))] \right\},
\end{equation}
where $p_{data}$ is the data-generating distribution. Both $\bm{G}$ and $\bm{D}$ are parameterized networks. 
%12-mars-25: 
%Note that, at fixed parameters of $\bm{G}$, the optimal \textit{discriminator} $\bm{D}^\ast$ has the following expression:
%\begin{equation}
%\bm{D}^\ast(\bm{x}) = \frac{p_{data}(\bm{x})}{p_{data}(\bm{x}) + p_{\bm{G}}(\bm{x})},
%\end{equation}
%with $p_{\bm{G}}(\bm{x}) = \pi(\bm{z})|\det J_{\bm{G}}|^{-1}$, where $J_{\bm{G}}$ is the Jacobian of the \textit{generator}. 
%This leads to a reformulation of the \texttt{min-max} problem as follows:
{\color{red}It can be shown that the \texttt{min-max} problem can be reformulated using the Jensen-Shannon metric to find the best generator $G$}
\begin{equation}
\min_G \left\{ \DKL{\left( p_{data} \left\| \frac{p_{data} + p_{\bm{G}}}{2}\right. \right)} + \DKL{\left(p_{\bm{G}} \left\| \frac{p_{data} + p_{\bm{G}}}{2} \right. \right)} \right\},
\end{equation}
yielding the optimal solution $p_{\bm{G}} = p_{data}$, which perfectly matches the dataset distribution.

@@

GAN architectures have been used for galaxy generation in conjunction with deblending and deconvolution tasks. For instance, \cite{Schawinski2017} and \cite{Hemmati_2022} employed vanilla-GANs, while \cite{Coccomini2021} utilized the \texttt{light-weight-GAN} that we use in our experiment too.
%
\subsection{Normalizing Flows}
\label{sec-NF}
%
Looking at Equation \ref{eq-VAE-ELBO}, one notices that the optimal solution is reached when $\DKL(p_{\bm{\phi}}(\bm{z}|\bm{x})\| p(\bm{z}|\bm{x})) = 0$, which is true if and only if $p_{\bm{\phi}}(\bm{z}|\bm{x}) = p(\bm{z}|\bm{x})$. However, this situation is generally unattainable because typical $p_{\bm{\phi}}$ parameterizations involve, for instance, independent Gaussian distributions. This limitation in parameterization is one of the drawbacks of VAEs. {\color{red} These considerations motivate the use of normalizing flows (NF) rethinking the problem of probability transport from $\pi(\bm{z})$ to $p_{data}(\bm{x})$. Introducing more flexibility in the family of distributions, we can hope that the true posterior is a member of such a family \citep{Tabak2010, Tabak2013a, Rezende2015}.}

%12-mars-25
%A flow $T$ is defined as a diffeomorphism (i.e., a bijector) between the data space and a latent space, such that:
%\begin{align}
%\bm{x} = T(\bm{z}) & \quad \Leftrightarrow \quad \bm{z} = T^{-1}(\bm{x}).
%\end{align}
%If the distribution of $\bm{z}$ is $\pi(\bm{z})$ (e.g., $\mathcal{N}(\bm{0},\bm{1})$), then the following relation applies:
%\begin{equation}
%p(\bm{x}) = \pi(\bm{z})|\det J_T(\bm{z})|^{-1} = \pi(T^{-1}(\bm{x}))|\det J_{T^{-1}}(\bm{x})|,
%\end{equation}
%where $J_T$ and $J_{T^{-1}}$ are the Jacobians of the transformations $T$ and $T^{-1}$, respectively. Unlike VAEs and GANs, the dimensionalities of the latent and data spaces are the same by definition in flow-based models.

{\color{red}
A flow $T$ is generally constructed as a composition of several individual diffeomorphism (i.e., a bijector)  $\{T_i\}_{i<n}$ such that:
\begin{align}
T = T_1 \circ T_2 \circ \dots \circ T_n & \quad \Leftrightarrow \quad 
T^{-1} = T_n^{-1} \circ T_{n-1}^{-1} \circ \dots \circ T_1^{-1}.
\end{align}
Unlike VAEs and GANs, the dimensionalities of the latent and data spaces are the same by definition in flow-based models.
Figure \ref{fig-normflow} illustrates a schematic view of the \textit{forward} or \textit{generative} direction and the \textit{backward/reverse} or \textit{training} direction. If we denote $\bm{z}_i = T_i(\bm{z}_{i-1})$ for all $i < n$ (using $\bm{z}_0 = \bm{z}$ and $\bm{z}_n = \bm{x}$), then:
\begin{equation}
\log |\det J_T| = \sum_{i=0}^{n-1} \log |\det J_{T_i}(\bm{z}_{i-1})|.
\label{eq-flow-jacob}
\end{equation}
}

@@

Let the flow be parametrized by a vector $\bm{\theta}$, leading to a parametrized flow model $p_{\bm{\theta}}(\bm{x})$. To optimize the flow, we consider the $\DKL$ divergence between $p_{\bm{\theta}}(\bm{x})$ and the true $p(\bm{x})$, resulting in\footnote{For simplicity, we assume $\pi(\bm{z}) = \mathcal{N}(\bm{0}, \bm{1})$ without introducing additional unknown parameters.}:
\begin{align}
\mathcal{L}(\bm{\theta}) &= \DKL(p(\bm{x})\|p_{\bm{\theta}}(\bm{x})) 
= - \Esp_{\bm{x}\sim p(\bm{x})}[-\log p_{\bm{\theta}}(\bm{x})] + \mathrm{const.} \nn \\
&= - \Esp_{\bm{x}\sim p(\bm{x})}[\log \pi(T_{\bm{\theta}}^{-1}(\bm{x})) + \log |\det J_{T_{\bm{\theta}}^{-1}}(\bm{x})|] + \mathrm{const.}
\label{eq-NF-loss-1}
\end{align}
{\color{red} Using this method, we directly address the convergence of $p_\theta$ towards $p$ (i.e., $p_{\text{data}}$) using the Kullback-Leibler metric, while avoiding the GAN issue related to the \texttt{min-max} Nash equilibrium.}

%In practice, computing the expectation $\Esp_{\bm{x} \sim p(\bm{x})}$ 
%is not feasible and is replaced by a Monte Carlo approach using the data samples $\{\bm{x}^{(i)}\}_{i<N}$. 
{\color{red} In practice, computing the expectation $\Esp_{\bm{x} \sim p(\bm{x})}$ 
is performed by a Monte Carlo approach using the data samples $\{\bm{x}^{(i)}\}_{i<N}$.} 
Dropping the constant term, the loss function becomes:
\begin{equation}
\mathcal{L}(\bm{\theta}) = -\frac{1}{N} \sum_{i=0}^{N-1} \left\{ \log \pi(T_{\bm{\theta}}^{-1}(\bm{x}^{(i)})) + \log |\det J_{T_{\bm{\theta}}^{-1}}(\bm{x}^{(i)})| \right\}.
\label{eq-NF-loss-2}
\end{equation}

The computation of the determinant of the Jacobian and its gradient can be challenging. However, it becomes more straightforward if the Jacobian matrices $\{J_{T_i}^{-1}\}_{i<n}$ (Eq.~\ref{eq-flow-jacob}) are triangular. To minimize $\mathcal{L}(\bm{\theta})$ and obtain the best $\bm{\theta}$, we require $T_\theta^{-1}$ as well as its Jacobian and gradients. Additionally, $T_\theta$ is needed to generate new samples $\bm{x}$ from latent samples $\bm{z} \sim \pi(\bm{z})$. Various architectures of normalizing flows, such as \textit{autoregressive flows}, enable such operations, as reviewed in \cite{Papamakarios2021}. {\color{red} We provide here some details, without aiming to be exhaustive, to clarify the design of certain architectures used in the literature.
}

The general schema of \textit{autoregressive flows} is as follows: let $\bm{x}, \bm{z} \in \mathbb{R}^d$, and denote $\bm{x} = (x_1, x_2, \dots, x_d)$ and $\bm{z} = (z_1, z_2, \dots, z_d)$. Then $\forall i$:
\begin{equation}
x_i = T(z_i; c_i) \quad \mathrm{with} \quad c_i = C_i(z_1, \dots, z_{i-1}) \triangleq C_i(z_{<i}),
\end{equation}
{\color{red}where $T$ knows as the \textit{transformer}\footnote{This term should not be confused with transformer networks used in deep learning architectures based on multi-head attention mechanisms \citep{Vaswani2017}.} is a strictly increasing monotonic transformation, and $C_i$ is known as the \textit{conditioner}.}

%12-Mars-25
%The properties of such flows ensure the invertibility of the \textit{transformer} and the triangular structure of the Jacobian:
%\begin{equation}
%J_T = \left[ \frac{\partial x_i}{\partial z_j} \right] = 
%\begin{cases}
%0 & \mathrm{for}\ i > j, \\
%\frac{\partial x_i}{\partial z_i}(z_i; c_i) & \mathrm{for}\ i = j.
%\end{cases}
%\end{equation}
%This leads to:
%\begin{equation}
%\log |\det J_T| = \sum_{i=1}^d \log \left| \frac{\partial x_i}{\partial z_i}(z_i; c_i) \right|.
%\end{equation}
%Different types of transformers and conditioners yield distinct architectures.

%12-mars-25 Among the transformers, the class of affine transformations is  perhaps the simplest. 
{\color{red} Among the transformers, the class of affine transformations is simple and widely used.} These are defined as:
\begin{equation}
T(z_i; c_i) = e^{\alpha_i} z_i + \beta_i, \quad \mathrm{with} \quad c_i = (\alpha_i(z_{<i}), \beta_i(z_{<i})).
\label{eq-affine-coupling}
\end{equation}
%Using matrix notation, this can be expressed as $T(\bm{z}) = \exp(\bm{\alpha}) \odot \bm{z} + \bm{\beta}$, where $\odot$ denotes the Hadamard product. 
The invertibility of $T$ is guaranteed by the exponential function, and the Jacobian determinant simplifies to the $e^{\alpha_i}$ factor. Affine transformers are used in works such as \citep{DinhKB14, Papamakarios2017a, DinhSB17, Kingma2018}. 
Furthermore, \textit{spline}-based flows have gained traction, as exemplified in \cite{Crenshaw_2024}, where the \texttt{PZFlow} code is applied to model galaxy photometric redshift posterior distributions.

Concerning the conditioners $C_i$, in principle, they can be any functions (or models) that take $z_{<i}$ as input and output $c_i$. However, practical implementations are driven by computational cost. A particular approach is the \textit{coupling layer}, implemented in \texttt{Real NVP}
%12-mars-25\footnote{n.b stands for \textit{real-valued non-volume preserving} transformation.} 
as proposed in \citep{DinhKB14, DinhSB17}. 
%12-Mars-25
%For an affine transformation $T$ (Eq.~\ref{eq-affine-coupling}) with $s < d$ (a common choice is $s = d/2$), the forward pass is defined as:
%\begin{equation}
%(\mathrm{NVP}) \quad \begin{cases}
%x_{\leq s} &= z_{\leq s}, \\
%x_{s+1,\dots,d} &= z_{s+1,\dots,d} \odot \exp(\bm{\alpha}(z_{\leq s})) + \bm{\beta}(z_{\leq s}),
%\end{cases}
%\label{eq-NVP_forward}
%\end{equation}
%where $\bm{\alpha}(z_{\leq s})$ and $\bm{\beta}(z_{\leq s})$ are functions of the first $s$ components of $\bm{z}$.
%
%The backward pass, involving the inverse transformation, has the same computational complexity—unlike other conditioners such as the masked schema used in \citep{Germain2015}—and is given by:
%\begin{equation}
%(\mathrm{NVP}^{-1}) \quad \begin{cases}
%z_{\leq s} &= x_{\leq s}, \\
%z_{s+1,\dots,d} &= (x_{s+1,\dots,d} - \bm{\beta}(x_{\leq s})) \odot \exp(-\bm{\alpha}(x_{\leq s})).
%\end{cases}
%\label{eq-NVP_backward}
%\end{equation}
%
%The Jacobian matrix in this schema is triangular, with the diagonal given by:
%\begin{equation}
%\diag(J_T) = (\bm{1}_s, \diag(\exp(\bm{\alpha}(z_{\leq s})))),
%\end{equation}
%where $\bm{1}_s$ is an $s \times s$ identity matrix. This leads to $d-s$ non-unity diagonal terms, reflecting the \textit{non-volume conservation} property of the transformation (hence the "NVP" notation).
%The structure of the transformation (e.g., Equation~\ref{eq-NVP_forward}) involves a single split of the $\bm{z}$ vector coordinates into two parts. The first part, $z_{\leq s}$, is left untransformed, while the transformation of the second part, $z_{>s}$, depends only on $z_{\leq s}$. Notably, the conditioner functions $\bm{\alpha}$ and $\bm{\beta}$ depend exclusively on $z_{\leq s}$ for both the forward and backward transformations. This single-split schema is illustrated in Figure~\ref{fig-nvp}.
{\color{red} In this context, the structure of the transformation involves a split of the $\bm{z}$ vector coordinates into two parts. The first part, $z_{\leq s}$ with $s=d/2$, is left untransformed (eg. $T=Id$), while the transformation of the second part (eg. affine), $z_{>s}$, depends only on $z_{\leq s}$. Notably, the conditioner functions $\bm{\alpha}$ and $\bm{\beta}$ depend exclusively on $z_{\leq s}$ for both the forward and backward transformations. 
}

%%12-mars-25
%\begin{figure}
%\centering
%\includegraphics[width=\linewidth]{fig-nvp.pdf}
%\includegraphics[width=\linewidth]{fig-nvp-inverse.pdf}
%\caption{Affine coupling layer schema of forward and backward directions (Equations~\ref{eq-NVP_forward},\ref{eq-NVP_backward}) with a single split mechanism (figure inspired by \protect\cite{Papamakarios2021}).}
%\label{fig-nvp}
%\end{figure}
%%
This splitting schema, implemented in a so-called \textit{coupling layer}, makes the flow-based model computationally efficient. By using deep neural networks, one can achieve complex $(\bm{\alpha},\bm{\beta})$ functions. Designers of complete architectures can compose different \textit{coupling layers}, alternating the order of $\bm{z}$ elements between layers to ensure that all elements are transformed by the end of the sequence of operations. A generalization of this permutation can be achieved using a $1\times 1$ convolution layer.

\cite{DinhSB17} further improved the modeling by introducing a \textit{multi-scale} approach, which is too extensive to detail here. \cite{Kingma2018} utilized these extensions to build the \texttt{Glow} model, which is employed in our experiments. For completeness, the schema of this model is depicted in Figure~\ref{fig-Glow-archi}.
%12-mars-25
% bored from Figure 2 of \cite{Kingma2018}. 
In the \texttt{Glow} model, the prior is learned during the optimization of the flows, but to generate new samples, one still uses $\bm{z} \sim \mathcal{N}(\bm{0}, \bm{1})$, where $\bm{z}$ comprises tensors corresponding to the number of levels $L$.

@@

%Using the presentation framework of 
%12-mars-26 pas la bonne reference \cite{kadkhodaie2024generalization}
{\color{red}Using the presentation framework of \cite{guth2022wavelet}}, the transport process in the \textit{score-based diffusion} algorithm is modeled by the Ornstein-Uhlenbeck equation \citep{Uhlenbeck1930}. The \textit{forward} process follows this stochastic differential equation:
\begin{equation}
d\bm{x}_t = -\bm{x}_t \, dt + \sqrt{2}\, dB_t \quad t\in[0,T],
\label{eq-Ornstein-cont}
\end{equation}
where $dB_t$ represents a Wiener process (Brownian motion) and $T$ is a finite endpoint assumed to be sufficiently large. The solution to this forward process is:
\begin{equation}
\bm{x}_t = \bm{x}_0 e^{-t} + \underbrace{(1-e^{-2t})^{1/2}}_{\sigma_t} \bm{z}, \qquad \bm{z} \overset{\text{iid}}{\sim} \mathcal{N}(\bm{0},\bm{1}).
\end{equation}
The corresponding probability density function is expressed as:
\begin{equation}
p_t(\bm{x}_t) = \int p_t(\bm{x}_t,\bm{x}_0) \, d\bm{x}_0 = \int p_t(\bm{x}_t|\bm{x}_0) p_0(\bm{x}_0) \, dx_0,
\label{eq-smoothing-p0-gauss}
\end{equation}
where, based on the Fokker-Planck equation for the Ornstein-Uhlenbeck process:
\begin{equation}
p_t(\bm{x}_t|\bm{x}_0) = \frac{1}{(2\pi \sigma_t^2)^{d/2}} \exp\left\{-\frac{\|\bm{x}_t-\bm{x}_0 e^{-t}\|^2}{2\sigma_t^2}\right\}.
\end{equation}
Thus, $p_t(\bm{x}_t)$ is the convolution of the initial distribution $p_0(\bm{x}_0)$ with a Gaussian of variance $\sigma_t^2$. This gradually blurs the distribution, leading $p_t(\bm{x}_t)$ to converge towards the normal distribution $\mathcal{N}(\bm{0},\bm{1})$.

@@

{\color{red}The central challenge in generation lies in estimating the \textit{score}, the key ingredient of the generation process.} Two main approaches have been proposed: (1) deriving a model for $p_t$ that incorporates regularity patterns and correlations to learn the \textit{score} via \textit{score matching} \citep{hyvarinen2005a}, or (2) employing a \textit{denoising} neural network, as introduced by \cite{Bengio2013}, further developed in \cite{Sohl-Dickstein2015,Ho2020}, and recently examined from a mathematical perspective in \cite{kadkhodaie2024generalization}. The first approach has been utilized in works such as \cite{Guth2022b,Lempereur2024} with Gibbs energy parameterization. The second has found applications in astrophysics, including galaxy image generation \citep{smith2021}, generation of 21~cm luminosity temperature maps \citep{Zhao2023}, super-resolution of large-scale cosmic structures \citep{Schanz2023}, and Bayesian posterior sampling for weak lensing mass-mapping problems \citep{Remy2023}.
